---
title: "Analyze of events on United States"
author: "Massimiliano Pinato"
date: "20 settembre 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis


# Data Processing
```{r read-data,echo=TRUE}
library(dplyr) # I use that for manipulate data
library(data.table) #I use that for faster read csv file
library(lubridate) #I use that for manipulate date/time
library(ggplot2) #I use that for plot
library(knitr) #I use that for knitr function like 'kable' for example
dfStorm <- fread("repdata-data-StormData.csv") #read file
```
I use the "FATALITIES" variable to determine the event most dangerouse for human healt.
```{r groupby-fatal,echo=TRUE}
dfStorm.overall <- dfStorm %>% 
  group_by(EVTYPE) %>% #group by event type
  summarize(FAT = sum(FATALITIES)) #calculate total of fatalities for each event
summary(dfStorm.overall) #I want to view how are the events
```
I see that the 3rd quarter is still with zero value. It means that only a bit of events have fatalities.
I try excluding events without fatalities.
```{r exclude-zero-fat,echo=TRUE}
dfStorm.overall.zero <- dfStorm.overall %>% 
  filter(FAT > 0) 
summary(dfStorm.overall.zero) #I want to view how are the events
```
It seems better. Now about 40 events are over 19.75 fatalities. I still want to see how are those 40 events.
```{r more-than-20,echo=TRUE}
dfStorm.overall.20 <- dfStorm.overall %>% 
  filter(FAT > 19.75) 
summary(dfStorm.overall.20) #I want to view how are the events
```

About 10 of these have more than 205. Now I want so know how much fatalities have those 205 respect al the others.

```{r analyze-205,echo=TRUE}
dfStorm.overall.mt205 <- dfStorm.overall %>% #events with more than 205 fatalities
  filter(FAT > 205) 
dfStorm.overall.lt205 <- dfStorm.overall %>% #events with less than 205 fatalities
  filter(FAT <= 205) 
#calculate total and event numbers of 'more than 205 fatalities'
fat.mt205 <- sum(dfStorm.overall.mt205$FAT) 
fat.mt205.cnt <- length(dfStorm.overall.mt205$EVTYPE)
#calculate total and event numbers of 'less than 205 fatalities'
fat.lt205 <- sum(dfStorm.overall.lt205$FAT)
fat.lt205.cnt <- length(dfStorm.overall.lt205$EVTYPE)
#prepare a dataframe in order to better present the data
df205 <- as.data.frame(
    rbind( 
      c("More than 205 fatalities",fat.mt205,fat.mt205.cnt),
      c("Less than 205 fatalities",fat.lt205,fat.lt205.cnt) 
    )
)
names(df205) <- c("GROUP","FATALITIES","EVTYPE.NUMBER")
```

Here the results. I see that the most of the fatalities (over 81%) is concentrate in a little part of events (only 11, about 1% of the overall events).

```{r group205-table,echo=TRUE}
library(knitr)
kable(df205)
```

Now I can see a bar plot of those events.

```{r group-mt205-plot,echo=TRUE}
g <- ggplot(dfStorm.overall.mt205,aes(x=reorder(EVTYPE,FAT),y=FAT))
g <- g + geom_bar(stat="identity",size=10)
g <- g + geom_text(aes(label=EVTYPE),vjust=-0.2)
g
```

# Results 

I've extracted the most dangerous events.
I've seen that 1% (about 10 events) of events produced the 81% of fatalities on overall events in the US.

```{r res-view-mt205-table,echo=TRUE}
dfStorm.overall.mt205vsOthers <- dfStorm.overall.mt205 %>% arrange(desc(FAT))
dfStorm.overall.mt205vsOthers <- rbind(dfStorm.overall.mt205vsOthers,
            data.frame(EVTYPE =  "--- THE OTHERS ---" ,
                       FAT = sum(dfStorm.overall.lt205$FAT)))
kable(dfStorm.overall.mt205vsOthers)
```

On thats events, Tornado is actually the most dangerous for the human health with more than 5000 fatalities.

```{r res-view-mt205-plot,echo=TRUE}
dfStorm.overall.mt205 <- dfStorm.overall.mt205 %>% 
  mutate(MOSTDANGEROUS = ifelse(EVTYPE == "TORNADO","MOST","LESS"))
g <- ggplot(dfStorm.overall.mt205,aes(x=reorder(EVTYPE,FAT),y=FAT,fill=MOSTDANGEROUS))
g <- g + geom_bar(stat="identity",size=10)
g <- g + geom_text(aes(label=EVTYPE),vjust=-0.2)
g
```
