---
title: "Analyze of events on United States"
author: "Massimiliano Pinato"
date: "20 settembre 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis


# Data Processing
We primarly read he CSV file.
```{r library-readfile,echo=TRUE}
library(dplyr) # I use that for manipulate data
library(data.table) #I use that for faster read csv file
library(lubridate) #I use that for manipulate date/time
library(ggplot2) #I use that for plot
library(knitr) #I use that for knitr function like 'kable' for example
dfStorm <- fread("repdata-data-StormData.csv") #read file
```

##Most Dangerous Events
For the dangerous events i prepare two dataset. One for FATALITIES and one for INJURIES, in a indipendent manner.
For each scenario I prepare the dataset of top 10 events by measure and a barplot to show graphically the result.
```{r read-data,echo=TRUE}
#Preparation of FATALITIES scenario
#Dataset for both totals
dfStorm.overall <- dfStorm %>%
  group_by(EVTYPE) %>% #grouping evtype
  summarize(FAT = sum(FATALITIES), INJ = sum(INJURIES)) #create totals
#Dataset of fatalities
dfStorm.overall.top10FAT <- dfStorm.overall %>% 
  top_n(10,FAT) %>% #select top 10 events by fatalities
  mutate(EVTYPE.ABR = abbreviate(EVTYPE,minlength=3)) #abbreviate for better graph
#Barplot of fatalities
g.fat <- ggplot(dfStorm.overall.top10FAT,aes(x=reorder(EVTYPE.ABR,FAT),y=FAT)) + 
  geom_bar(stat="identity") + 
  xlab("Event Types") + ylab("Fatalities") + labs(title="Top 10 events for fatalities")
#Dataset of injuries
dfStorm.overall.top10INJ <- dfStorm.overall %>% 
  top_n(10,INJ) %>% #top 10 events for injurius
  mutate(EVTYPE.ABR = abbreviate(EVTYPE,minlength=3)) #abbreviate for better presentation
#Barplot of injuries
g.inj <- ggplot(dfStorm.overall.top10INJ,aes(x=reorder(EVTYPE.ABR,INJ),y=INJ)) +
  geom_bar(stat="identity") + 
  xlab("Event Types") + ylab("Injurious") + labs(title="Top 10 events for injurious")
```

##Greatest economic consequences



# Results 

##Most Dangerous Events
According to data dictionary, dangerous is rappresented by fatalities and injourious.
We have extracted top 10 events by these measures.
For fatalities, we have the following situation.
```{r show-fat,echo=TRUE}
kable(dfStorm.overall.top10FAT)
g.fat
```
The event "TORNADO" is the most dangerous for fatalities.
In the case of injourious, we have the following situation.
```{r show-inj,echo=TRUE}
kable(dfStorm.overall.top10INJ)
g.inj
```
Even in this case, "TORNADO" is the most dangerous.
So the conclusion is that "TORNADO" is always the most dangerous event for people.

##Greatest economic consequences
